Summary of the things that I learned

### 2024-10-22 700-formalization.tex
Process feedback about introduction, conclusion, bibliography and citations

\section{Some Numbers}
The code for this project was written over the course of about 18 months, spread over \href{https://github.com/UniMath/UniMath/pulls?q=is%3Apr+author%3Aarnoudvanderleer+is%3Aclosed}{20 pull requests}: 13 with content about Hyland's paper, adding 23.361 and removing 7.620 lines of code, and 7 with only some missing category theoretical preliminaries, adding 3.291 and removing 816 lines.

@@

The formalized parts of Hyland's paper are collected in a package in the library called \textit{algebraic theories}. This package now consists of over 14.000 lines of code, and contains

@@

Of course, it was possible to get rid of the objects altogether, and just define things in terms of the objects and the morphisms of the categories. However, in practice this causes problems because of \textit{coercions}. For example, if $ X $ is an algebraic theory, mathematicians like to use the name $ X $ also to denote the sequence of sets $ X_n $. Under the hood, this uses a coercion, which allows one to use the same symbol to denote both the entire object, or a part of it, depending on the context. However, when working with displayed categories, the category of algebraic theories is displayed over the category of algebraic theory data, which is displayed over the category of sequences of sets. It turns out that in coq, coercions on categories do not compose very well: If $ X $ is an algebraic theory, and we have coercions from algebraic theories to algebraic theory data, and from algebraic theory data to sequences of sets, we can still only use $ X $ to denote the algebraic theory data, and not the sequence of sets.

@@

\section{Duplication in Definitions}
One of the adages in software engineering is `DRY': ``Don't Repeat Yourself'': Long expressions and blocks of code that occur multiple times throughout the program should usually be abstracted into a separate function. One of the reasons for this is that it is easier to change code in this function, than to change every instance of the repeated expression or block of code. When writing mathematics, such functions are usually called `lemmas'. However, in this project there was also another example of duplication, which occurred in statements of definitions and lemmas. For example, in the definition of the displayed category of algebraic theories, the displayed object type over a sequence of sets $ T_n $ is
\[ \left(\prod_n \prod_{i : \{ 1, \dots, n \}} T_n \right) \times \left( \prod_{m, n} T_m \times T_n^m \to T_n \right), \]
corresponding to the variables and the substitution. Then, the constructor of an algebraic theory takes arguments
\[ v : \prod_n \prod_{i : \{ 1, \dots, n \}} T_n \quad \text{and} \quad s : \prod_{m, n} T_m \times T_n^m \to T_n. \]
Also, given an algebraic theory, we have accessors:
\[ x_{n, i} : T_n \quad \text{and} \quad \bullet_{m, n} : (\prod_{m, n} T_m \times T_n^m \to T_n) \]
Lastly, when we define a new algebraic theory, we need to provide terms of these types again.

@@

This indeed reduces the amount of duplication in the code, and makes it slightly easier to write some definitions, because one does not have to remember and type the exact formulation of every axiom. However, there is no free lunch here: The axioms are not immediately unfolded when they occur. Therefore, if one uses the constructor for an algebraic theory, the goals become \texttt{var\_ax} and \texttt{subst\_ax}, and these have to be unfolded to see what they mean. Also, when coq is asked to state the property of an algebra morphism, it responds with \texttt{mor\_action\_ax}, which is not very informative. This is not a big problem, per se, but it adds some friction when formalizing. This friction could be reduced a lot if coq would have some sort of macros, which would immediately be unfolded upon use, and would never be printed.

@@

Also, this project added a lemma to the library stating that in any category $ C $, $ \prod_{i : \emptyset} X_i $ (for $ X : \emptyset \to T $ a family of objects) is given by the terminal object. However, the repeated binary product construction needed $ \prod_{i : \{ i \mid 1 \leq i \leq 0 \}} X_i $, and $ \emptyset $ is not definitionally equal to $ \{ i \mid 1 \leq i \leq 0 \} $. Initially, this was solved by transporting along the equality $ \emptyset = \{ i \mid 1 \leq i \leq 0 \} $. However, much further down the line, some proof involved an element of this product over $ \{ i \mid 1 \leq i \leq 0 \} $, and because of the transport, coq was not able to see that this product was just $ \{ \star \} $. Therefore, the statement of the lemma about the empty product was changed to ``For a type $ I $ and a family $ X : I \to C $, if we have a function $ f : I \to \emptyset $, the terminal object gives the product $ \prod_{i : I} X_i $''. Note that having the function $ f $ is equivalent to $ I $ being equivalent (and therefore equal) to the empty type. This change in the statement did provide the right amount of generality to make computation down the line much smoother.

@@

The second occurrence of quotients was in the proof of Corollary \ref{cor:karoubi-presheaf}, or the lemma that it is based on. Initially, the statement of the lemma was only about $ \SET $, stating that $ \iota_C \bullet - : [\bar C, \SET] \to [C, \SET] $ was an adjoint equivalence. The proof used coequalizers of functors, like in Lemma \ref{lem:karoubi-equivalence}. Note that coequalizers of set-valued functions are given pointwise by coequalizers in $ \SET $, which are formalized using quotients. Because of this, when simplifying, coq usually reduced the goal to elementary statements about quotients, which somewhat obfuscated the relatively simple nature of the construction. Luckily, the construction could be made agnostic to the specific implementation of coequalizers, by introducing a variable
\[ \mathtt{H : Coequalizers HSET} \]
and using this instead of our known implementation. Since the proofs did not use any elements of the sets in question, it was then trivial to generalize to any category with coequalizers. It turned out that the construction was a special version of left Kan extension, which uses general colimits, and using the existing definition of Kan extension, the proof was made considerably shorter.

@@

\section{The Formalization of the \texorpdfstring{$ \lambda $}{lambda}-calculus} \label{sec:axiomatic-lambda-calculus}
One of the important classes of objects in this thesis is the class of $ \Lambda $-algebras: algebras for the initial $ \lambda $-theory $ \Lambda $, corresponding to the `pure' $ \lambda $-calculus. Therefore, much of this project required the pure $ \lambda $-calculus to be formalized. Now, in this thesis, it is defined as a quotient of an inductive type. However, in the UniMath library, the use of inductive types (other than a handful of basic ones like \texttt{empty}, \texttt{bool}, \texttt{nat} and \texttt{coprod}) is prohibited
\footnote{This is because allowing inductive types requires a much larger trusted codebase of the proof assistant, and because most of mathematics can be formalized with $ \sum $-types or using initial algebras of functors, instead of inductive and record types.}. Also, as we saw in section \ref{sec:quotients}, working with quotients can be somewhat complicated.

The inductive type approach has a variation, which instead of a quotient uses a higher inductive type to force the required elements to be equal. Unfortunately, even if the use of inductive types was allowed, coq does not support higher inductive types.

Instead, a different, more axiomatic approach was taken. A class in group theory often starts with laying out a couple of axioms, like: ``We have a set $ G $, with a binary operation $ b $, a unary operation $ u $ and an element $ e $. $ b $ is associative, $ u $ is a left and right inverse for $ b $ and $ e $ is a unit for $ b $.'' Afterwards, this structure is usually bundled into the declaration ``Let $ G $ be a group.'' In the same spirit, every section in the formalization that needs the pure $ \lambda $-calculus, starts with ``Let $ L $ be the pure $ \lambda $-calculus with $ \beta $-equality.'' To this end, there is a definition \texttt{lambda\_calculus}, consisting of
\begin{itemize}
  \item A sequence of sets $ (L_n)_n $;
  \item Variables, app, abs, subst:
    \[ \mathtt{var}_{n, i} : L_n, \quad \mathtt{app}_n: L_n \times L_n \to L_n, \quad \mathtt{abs}_n : L_{S n} \to L_n \quad \text{and} \quad \mathtt{subst}_{m, n}: L_m \times L_n^m \to L_n; \]
  \item A couple of identities about the interactions between the constructors. In particular, $ \beta $-equality:
    \[
      \mathtt{subst}(\mathtt{var}_i, t) = t_i, \quad
      \dots, \quad
      \mathtt{app}(\mathtt{abs}(s), t) = \mathtt{subst}(s, (\mathtt{var}_i)_i + (t));
    \]
  \item The induction principle, which coincides with the induction principle that a higher inductive type would have: Given, for all $ t : L_n $, a type $ A_{n, t} $, it is possible to construct, for every $ t : L_n $, an element $ f(t) : A_{n, t} $ by just giving elements and functions, corresponding to the constructors
    \begin{align*}
      f_{\mathtt{var}(n, i)} &:A_{\mathtt{var}_{n, i}}\\
      f_{\mathtt{app}(s, t)} &:A_s \times A_t \to A_{\mathtt{app}(s, t)}\\
      f_{\mathtt{abs}(t)} &:A_t \to A_{\mathtt{abs}(t)}\\
      f_{\mathtt{subst}(s, t)} &:A_s \times A_{t_1} \times \dots \times A_{t_n} \to A_{\mathtt{subst}(s, t)}\\
    \end{align*}
    and by showing that they are compatible with the identities on $ L $. For example, for $ \beta $-equality, this is equality between
    \[ f_{\mathtt{app}}(f_{\mathtt{abs}}(s), t): A_{\mathtt{app}(\mathtt{abs}(s), t)} \quad \text{and} \quad f_{\mathtt{subst}}(s, f_{\mathtt{var}_1}, \dots, f_{\mathtt{var}_n}, t): A_{\mathtt{subst}(s, (\mathtt{var}_i)_i + (t))}. \]
    Note that the left and right hand side live in different types, so the equality can only be stated using a transport over $ \beta $-equality.
  \item A couple of identities about the interaction between induction and the constructors: if $ f $ is defined using the induction principle as above,
  \begin{align*}
    f(\mathtt{var}_i) &= f_{\mathtt{var}_i}\\
    f(\mathtt{app}(s, t)) &= f_{\mathtt{app}(s, t)}(f(s), f(t))\\
    f(\mathtt{abs}(t)) &= f_{\mathtt{abs}(t)}(f(t))\\
    f(\mathtt{subst}(s, t)) &= f_{\mathtt{subst}(s, t)}(f(s), (f(t_i))_i).
  \end{align*}
\end{itemize}

@@

Therefore, along with the original proof of Scott's representation theorem, a new version of the tactic, called \texttt{propagate\_subst} has been added. It recursively traverses the $ \lambda $-term in the left-hand side of the goal, checking whether the term matches a form that can be rewritten into something else. It performs the possible rewrites, and also prints these rewrite statements which can replace it. For example, if the goal is
\begin{lstlisting}
(λ'm, (inflate a (inflate b var (stnweq (inr tt))))) • c =
(λ'n, (inflate (a • c) (inflate (b • c) var (stnweq (inr tt)))))
\end{lstlisting}
a call to \texttt{repeat reduce\_lambda} would take about a second, but a call to \texttt{propagate\_subst ()} runs in about $ \frac 1 3 $ of a second and prints
\begin{lstlisting}
refine '(subst_abs _ _ _ @ _).
refine '(_ @ !maponpaths (λ x, (abs (app x _))) (inflate_subst _ _ _)).
refine '(_ @ !maponpaths (λ x, (abs (app _ (app x _)))) (inflate_subst _ _ _)).
refine '(maponpaths (λ x, (abs x)) (subst_app _ _ _ _) @ _).
refine '(maponpaths (λ x, (abs (app x _))) (subst_inflate _ _ _) @ _).
refine '(maponpaths (λ x, (abs (app _ x))) (subst_app _ _ _ _) @ _).
refine '(maponpaths (λ x, (abs (app _ (app x _)))) (subst_inflate _ _ _) @ _).
refine '(maponpaths (λ x, (abs (app _ (app _ x)))) (var_subst _ _ _) @ _).
refine '(maponpaths (λ x, (abs (app _ (app _ x)))) (extend_tuple_inr _ _ _) @ _).
\end{lstlisting}
Replacing the call to \texttt{propagate\_subst} by these statements, results in the same rewrites, but these only take $ \frac{1}{25} $ of a second.

@@

It would be interesting to see whether parts of these tactics can be generalized. For example, whether it is possible to create an extensible tactic which shows that some type is a homotopy set, or a mere proposition, and prints the coq statements that can replace the tactic call. Or whether it would be possible to generalize \texttt{generate\_refine} to very quickly generate very fast and precise rewrite statements for large and complicated goals.

\section{The Learning Curve}
Let me conclude this chapter with a more personal note about formalization. In the past fifteen years, I have worked with a plethora of different programming languages, and I would like to think that in all this time, I have learned a great deal about programming. When I start working with a new language, it takes me a couple of minutes to figure things out, and within an hour, I can probably get a small program up and running, with some help from google. However, when I started working with coq and UniMath, I kind of had to start from scratch again. As it turns out, the naive or direct approach is often not the right one when working with a proof assistant, and of the code that I wrote in the first months, almost nothing remains. Even though every programming language has some sort of learning curve, formalizing in a proof assistant is especially unforgiving: You often have to change an old definition, or start over when you are halfway through a proof, because your current approach gets bogged down further and further with, for example, unintelligible transports. For every line of code that ends up in a pull request, many lines are written and erased again.

>>>>

\section{Some Numbers}
The code for this project was written over the course of about 18 months, spread over \href{https://github.com/UniMath/UniMath/pulls?q=is%3Apr+author%3Aarnoudvanderleer+is%3Aclosed}{20 pull requests}: 13 with content about Hyland's paper, adding 23\,361 and removing 7\,620 lines of code, and 7 with only some missing category theoretical preliminaries, adding 3\,291 and removing 816 lines.

@@

The formalized parts of Hyland's paper are collected in a package in the library called \textit{algebraic theories}. This package now consists of over 14\,000 lines of code, and contains:

@@

Of course, it was possible to get rid of the objects altogether, and just define things in terms of the objects and the morphisms of the categories. However, in practice this causes problems because of \textit{coercions}. For example, if $ X $ is an algebraic theory, mathematicians like to use the name $ X $ also to denote the sequence of sets $ X_n $. Under the hood, this uses a coercion, which allows one to use the same symbol to denote both the entire object, or a part of it, depending on the context. However, when working with displayed categories, the category of algebraic theories is displayed over the category of algebraic theory data, which is displayed over the category of sequences of sets. It turns out that in rocq, coercions on categories do not compose very well: If $ X $ is an algebraic theory, and we have coercions from algebraic theories to algebraic theory data, and from algebraic theory data to sequences of sets, we can still only use $ X $ to denote the algebraic theory data, and not the sequence of sets.

@@

\section{Duplication in Definitions}
One of the adages in software engineering is `DRY', which stands for ``Don't Repeat Yourself'': Long expressions and blocks of code that occur multiple times throughout the program should usually be abstracted into a separate function. One of the reasons for this is that it is easier to change code in this function, than to change every instance of the repeated expression or block of code. When writing mathematics, such functions are usually called `lemmas'. However, in this project there was also another example of duplication, which occurred in statements of definitions and lemmas. For example, in the definition of the displayed category of algebraic theories, the displayed object type over a sequence of sets $ T_n $ is
\[ \left(\prod_n \prod_{i : \{ 1, \dots, n \}} T_n \right) \times \left( \prod_{m, n} T_m \times T_n^m \to T_n \right), \]
corresponding to the variables and the substitution. Then, the constructor of an algebraic theory takes arguments
\[ v : \prod_n \prod_{i : \{ 1, \dots, n \}} T_n \quad \text{and} \quad s : \prod_{m, n} T_m \times T_n^m \to T_n. \]
Also, given an algebraic theory, we have accessors:
\[ x_{n, i} : T_n \quad \text{and} \quad \bullet_{m, n} : (\prod_{m, n} T_m \times T_n^m \to T_n) \]
Lastly, when we define a new algebraic theory, we need to provide terms of these types again.

@@

This indeed reduces the amount of duplication in the code, and makes it slightly easier to write some definitions, because one does not have to remember and type the exact formulation of every axiom. However, there is no free lunch here: The axioms are not immediately unfolded when they occur. Therefore, if one uses the constructor for an algebraic theory, the goals become \texttt{var\_ax} and \texttt{subst\_ax}, and these have to be unfolded to see what they mean. Also, when rocq is asked to state the property of an algebra morphism, it responds with \texttt{mor\_action\_ax}, which is not very informative. This is not a big problem, per se, but it adds some friction when formalizing. This friction could be reduced a lot if rocq would have some sort of macros, which would immediately be unfolded upon use, and would never be printed.

@@

Also, this project added a lemma to the library stating that in any category $ C $, $ \prod_{i : \emptyset} X_i $ (for $ X : \emptyset \to T $ a family of objects) is given by the terminal object. However, the repeated binary product construction needed $ \prod_{i : \{ i \mid 1 \leq i \leq 0 \}} X_i $, and $ \emptyset $ is not definitionally equal to $ \{ i \mid 1 \leq i \leq 0 \} $. Initially, this was solved by transporting along the equality $ \emptyset = \{ i \mid 1 \leq i \leq 0 \} $. However, much further down the line, some proof involved an element of this product over $ \{ i \mid 1 \leq i \leq 0 \} $, and because of the transport, rocq was not able to see that this product was just $ \{ \star \} $. Therefore, the statement of the lemma about the empty product was changed to ``For a type $ I $ and a family $ X : I \to C $, if we have a function $ f : I \to \emptyset $, the terminal object gives the product $ \prod_{i : I} X_i $''. Note that having the function $ f $ is equivalent to $ I $ being equivalent (and therefore equal) to the empty type. This change in the statement did provide the right amount of generality to make computation down the line much smoother.

@@

The second occurrence of quotients was in the proof of Corollary \ref{cor:karoubi-presheaf}, or the lemma that it is based on. Initially, the statement of the lemma was only about $ \SET $, stating that $ \iota_C \bullet - : [\bar C, \SET] \to [C, \SET] $ was an adjoint equivalence. The proof used coequalizers of functors, like in Lemma \ref{lem:karoubi-equivalence}. Note that coequalizers of set-valued functions are given pointwise by coequalizers in $ \SET $, which are formalized using quotients. Because of this, when simplifying, rocq usually reduced the goal to elementary statements about quotients, which somewhat obfuscated the relatively simple nature of the construction. Luckily, the construction could be made agnostic to the specific implementation of coequalizers, by introducing a variable
\[ \mathtt{H : Coequalizers HSET} \]
and using this instead of our known implementation. Since the proofs did not use any elements of the sets in question, it was then trivial to generalize to any category with coequalizers. It turned out that the construction was a special version of left Kan extension, which uses general colimits, and using the existing definition of Kan extension, the proof was made considerably shorter.

@@

\section{The Formalization of the \texorpdfstring{$ \lambda $}{lambda}-calculus} \label{sec:axiomatic-lambda-calculus}
One of the important classes of objects in this thesis is the class of $ \Lambda $-algebras: algebras for the initial $ \lambda $-theory $ \Lambda $, corresponding to the `pure' $ \lambda $-calculus. Therefore, much of this project required the pure $ \lambda $-calculus to be formalized. Now, in this thesis, it is defined as a quotient of an inductive type. However, in the UniMath library, the use of inductive types (other than a handful of basic ones like \texttt{empty}, \texttt{bool}, \texttt{nat} and \texttt{coprod}) is prohibited\footnote{This is because allowing inductive types requires a much larger trusted codebase of the proof assistant, and because most of mathematics can be formalized with $ \sum $-types or using initial algebras of functors, instead of inductive and record types.}. Also, as we saw in section \ref{sec:quotients}, working with quotients can be somewhat complicated.

The inductive type approach has a variation, which instead of a quotient uses a higher inductive type to force the required elements to be equal. Unfortunately, even if the use of inductive types was allowed, rocq does not support higher inductive types.

Instead, a different, more axiomatic approach was taken. A class in group theory often starts with laying out a couple of axioms, like: ``We have a set $ G $, with a binary operation $ b $, a unary operation $ u $ and an element $ e $. $ b $ is associative, $ u $ is a left and right inverse for $ b $ and $ e $ is a unit for $ b $.'' Afterwards, this structure is usually bundled into the declaration ``Let $ G $ be a group.'' In the same spirit, every section in the formalization that needs the pure $ \lambda $-calculus, starts with ``Let $ L $ be the pure $ \lambda $-calculus with $ \beta $-equality.'' To this end, there is a definition \texttt{lambda\_calculus}, consisting of:
\begin{itemize}
  \item A sequence of sets $ (L_n)_n $;
  \item Variables, app, abs, subst:
    \[ \mathtt{var}_{n, i} : L_n, \quad \mathtt{app}_n: L_n \times L_n \to L_n, \quad \mathtt{abs}_n : L_{S n} \to L_n \quad \text{and} \quad \mathtt{subst}_{m, n}: L_m \times L_n^m \to L_n; \]
  \item A couple of identities about the interactions between the constructors. In particular, $ \beta $-equality:
    \[
      \mathtt{subst}(\mathtt{var}_i, t) = t_i, \quad
      \dots, \quad
      \mathtt{app}(\mathtt{abs}(s), t) = \mathtt{subst}(s, (\mathtt{var}_i)_i + (t));
    \]
  \item The induction principle, which coincides with the induction principle that a higher inductive type would have: Given, for all $ t : L_n $, a type $ A_{n, t} $, it is possible to construct, for every $ t : L_n $, an element $ f(t) : A_{n, t} $ by just giving elements and functions, corresponding to the constructors
    \begin{align*}
      f_{\mathtt{var}(n, i)} &:A_{\mathtt{var}_{n, i}}\\
      f_{\mathtt{app}(s, t)} &:A_s \times A_t \to A_{\mathtt{app}(s, t)}\\
      f_{\mathtt{abs}(t)} &:A_t \to A_{\mathtt{abs}(t)}\\
      f_{\mathtt{subst}(s, t)} &:A_s \times A_{t_1} \times \dots \times A_{t_n} \to A_{\mathtt{subst}(s, t)}\\
    \end{align*}
    and by showing that they are compatible with the identities on $ L $. For example, for $ \beta $-equality, this is equality between
    \[ f_{\mathtt{app}}(f_{\mathtt{abs}}(s), t): A_{\mathtt{app}(\mathtt{abs}(s), t)} \quad \text{and} \quad f_{\mathtt{subst}}(s, f_{\mathtt{var}_1}, \dots, f_{\mathtt{var}_n}, t): A_{\mathtt{subst}(s, (\mathtt{var}_i)_i + (t))}. \]
    Note that the left and right hand side live in different types, so the equality can only be stated using a transport over $ \beta $-equality.
  \item A couple of identities about the interaction between induction and the constructors: if $ f $ is defined using the induction principle as above,
  \begin{align*}
    f(\mathtt{var}_i) &= f_{\mathtt{var}_i}\\
    f(\mathtt{app}(s, t)) &= f_{\mathtt{app}(s, t)}(f(s), f(t))\\
    f(\mathtt{abs}(t)) &= f_{\mathtt{abs}(t)}(f(t))\\
    f(\mathtt{subst}(s, t)) &= f_{\mathtt{subst}(s, t)}(f(s), (f(t_i))_i).
  \end{align*}
\end{itemize}

@@

Therefore, along with the original proof of Scott's representation theorem, a new version of the tactic, called \texttt{propagate\_subst} has been added. It recursively traverses the $ \lambda $-term in the left-hand side of the goal, checking whether the term matches a form that can be rewritten into something else. It performs the possible rewrites, and also prints these rewrite statements which can replace it. For example, if the goal is
\begin{lstlisting}
(λ'm, (inflate a (inflate b var (stnweq (inr tt))))) • c =
(λ'n, (inflate (a • c) (inflate (b • c) var (stnweq (inr tt)))))
\end{lstlisting}
a call to \texttt{repeat reduce\_lambda} would take about a second, but a call to \texttt{propagate\_subst ()} runs in about $ 330 $ millseconds and prints
\begin{lstlisting}
refine '(subst_abs _ _ _ @ _).
refine '(_ @ !maponpaths (λ x, (abs (app x _))) (inflate_subst _ _ _)).
refine '(_ @ !maponpaths (λ x, (abs (app _ (app x _)))) (inflate_subst _ _ _)).
refine '(maponpaths (λ x, (abs x)) (subst_app _ _ _ _) @ _).
refine '(maponpaths (λ x, (abs (app x _))) (subst_inflate _ _ _) @ _).
refine '(maponpaths (λ x, (abs (app _ x))) (subst_app _ _ _ _) @ _).
refine '(maponpaths (λ x, (abs (app _ (app x _)))) (subst_inflate _ _ _) @ _).
refine '(maponpaths (λ x, (abs (app _ (app _ x)))) (var_subst _ _ _) @ _).
refine '(maponpaths (λ x, (abs (app _ (app _ x)))) (extend_tuple_inr _ _ _) @ _).
\end{lstlisting}
Replacing the call to \texttt{propagate\_subst} by these statements, results in the same rewrites, but these only take $ 40 $ milliseconds.

@@

It would be interesting to see whether parts of these tactics can be generalized. For example, whether it is possible to create an extensible tactic which shows that some type is a homotopy set, or a mere proposition, and prints the rocq statements that can replace the tactic call. Or whether it would be possible to generalize \texttt{generate\_refine} to very quickly generate very fast and precise rewrite statements for large and complicated goals.

\section{The Learning Curve}
Let me conclude this chapter with a more personal note about formalization. In the past fifteen years, I have worked with a plethora of different programming languages, and I would like to think that in all this time, I have learned a great deal about programming. When I start working with a new language, it takes me a couple of minutes to figure things out, and within an hour, I can probably get a small program up and running, with some help from google. However, when I started working with rocq and UniMath, I kind of had to start from scratch again. As it turns out, the naive or direct approach is often not the right one when working with a proof assistant, and of the code that I wrote in the first months, almost nothing remains. Even though every programming language has some sort of learning curve, formalizing in a proof assistant is especially unforgiving: You often have to change an old definition, or start over when you are halfway through a proof, because your current approach gets bogged down further and further with, for example, unintelligible transports. For every line of code that ends up in a pull request, many lines are written and erased again.

### 2024-10-22 800-conclusion.tex
Process feedback about introduction, conclusion, bibliography and citations

Also, since we saw that in univalent foundations, there are two nonequivalent definitions $ \bar \C $ and $ \hat \C $ for the category of retracts, it would be interesting to see how well Scott's and Taylor's proofs about $ \bar \C $, can be made to work on its Rezk completion $ \hat \C $. More generally, note that the Karoubi envelope is not specific to the material in this thesis. For example, one way to construct a cartesian closed category is by taking the Karoubi envelope of a `semi cartesian closed category', and the cartesian closed structure on $ \R $ in Scott's representation theorem can be viewed as a special case of this \autocite{hayashi-1985-semifunctors}. In another direction, the category of smooth manifolds can be constructed as the Karoubi envelope of the category $ \C $ of the open subsets of all Euclidean spaces, with smooth maps between them (see \autocite{Lawvere-Karoubi}, page 267). For such classical results about the Karoubi envelope, it would be interesting to study how they hold up in univalent foundations for the different choices $ \overline \C $ and $ \hat \C $ of `the Karoubi envelope'.

>>>>

Also, since we saw that in univalent foundations, there are two nonequivalent definitions $ \bar \C $ and $ \hat \C $ for the category of retracts, it would be interesting to see how well Scott's and Taylor's proofs about $ \bar \C $, can be made to work on its Rezk completion $ \hat \C $. More generally, note that the Karoubi envelope is not specific to the material in this thesis. For example, one way to construct a cartesian closed category is by taking the Karoubi envelope of a `semi cartesian closed category', and the cartesian closed structure on $ \R $ in Scott's representation theorem can be viewed as a special case of this \autocite{hayashi-1985-semifunctors}. In another direction, the category of smooth manifolds can be constructed as the Karoubi envelope of the category $ \C $ of the open subsets of all Euclidean spaces, with smooth maps between them \autocite[][267]{Lawvere-Karoubi}. For such classical results about the Karoubi envelope, it would be interesting to study how they hold up in univalent foundations for the different choices $ \overline \C $ and $ \hat \C $ of `the Karoubi envelope'.

### 2024-10-22 910-alternative-definitions.tex
Process feedback about introduction, conclusion, bibliography and citations

In this section, we will denote the finite set $ \llbracket n \rrbracket = \{ 1, \dots, n \} $.

@@

\begin{definition}
  A \iindex{cartesian operad} T is a functor $ T: F \to \SET $, together with an `identity' element $ \id{}: T(1) $ and for all $ m, n_1, \dots, n_m: \mathbb N $, a composition map
  \[ T(m) \times \prod_i T(n_i) \to T\left(\sum_i n_i\right), \]
  written $ (f, g_1, \dots, g_m) \mapsto f[g_1, \dots, g_m] $, satisfying some identity, associativity and naturality conditions (see \autocite{lambda-monoid}, Definition 2.1, for more details).
\end{definition}

@@

\begin{remark}
  Note that for an adjunction $ F \dashv G $, $ F \bullet G $ gives a monad. In the same way, there exists a notion of \iindex{relative adjunction}, from which we can obtain a relative monad (see \autocite{monads-endofunctors}, Theorem 2.10).
\end{remark}

@@

\begin{definition}
  Given a (skew-)monoidal category $ (C, \otimes, I) $, a \iindex{monoid} in this category is an object $ T : C $, together with a `multiplication' $ \mu: C(T \otimes T, T) $ and a `unit' morphism $ \eta: C(I, T) $ satisfying a couple of laws (see \autocite{MacLane}, Section III.6).
\end{definition}

>>>>

In this appendix, we will denote the finite set $ \llbracket n \rrbracket = \{ 1, \dots, n \} $.

@@

\begin{definition}
  A \iindex{cartesian operad} T is a functor $ T: F \to \SET $, together with an `identity' element $ \id{}: T(1) $ and for all $ m, n_1, \dots, n_m: \mathbb N $, a composition map
  \[ T(m) \times \prod_i T(n_i) \to T\left(\sum_i n_i\right), \]
  written $ (f, g_1, \dots, g_m) \mapsto f[g_1, \dots, g_m] $, satisfying some identity, associativity and naturality conditions. See \autocite[][Definition 2.1]{lambda-monoid} for more details.
\end{definition}

@@

\begin{remark}
  Note that for an adjunction $ F \dashv G $, $ F \bullet G $ gives a monad. In the same way, there exists a notion of \iindex{relative adjunction}, from which we can obtain a relative monad \autocite[][Theorem 2.10]{monads-endofunctors}.
\end{remark}

@@

\begin{definition}
  Given a (skew-)monoidal category $ (C, \otimes, I) $, a \iindex{monoid} in this category is an object $ T : C $, together with a `multiplication' $ \mu: C(T \otimes T, T) $ and a `unit' morphism $ \eta: C(I, T) $ satisfying a couple of laws \autocite[][Section III.6]{MacLane}.
\end{definition}
